\chapter{Conclusion}
\label{ch:conclusion}

The results in this dissertation comprise many detailed measurements of two-pion correlation functions in \pPb collisions at \pPbenergy with the ATLAS detector at the \lhc.
The size and shape of the particle source are presented as a function of event centrality, transverse momentum, rapidity, and azimuthal angle with respect to the second-order event plane.
These include the Lorentz invariant HBT radius \Rinv as well as the main 3D radii \Rout, \Rside, and \Rlong.
The cross-terms \Ros (coupling ``out'' and ``side'' components) and \Rol (coupling ``out'' and ``long'') are included in the azimuthally- and rapidity-dependent analyses, respectively.

The procedures developed in these results include some major technological improvements in the analysis procedure compared to previous measurements.
A data-driven technique is developed for constraining the contribution of jet fragmentation to the correlation function, which is a dominant systematic in small systems, particularly \pp and peripheral \pA collisions.
This approach reduces the number of assumptions necessary for describing this background.
A sampling method is also developed to correct the azimuthally-dependent correlation functions for the event plane resolution, addressing an oversight made in the literature up to this point.

The HBT radii, representing the size of the source's region of homogeneity along the outwards (along \kt), sideways (other transverse), and longitudinal axes, are significantly larger in central collisions.
In central collisions, the radii exhibit a strong decrease with rising \kt, which is indicative of collective expansion.
This trend is significantly diminished in peripheral collisions, where within uncertainties there is no significant slope of the radii with respect to \kt.
Hydrodynamics may therefore be an appropriate description of central \pPb collisions and yet fail in peripheral ones.

Within a \kt interval, the source radii increase monotonically from peripheral to central collisions.
At lower \kt the slope of this increase is larger and the radii are proportional to \(\avgdNdeta^{1/3}\).
The radii are evaluated as a function of the rapidity-dependent multiplicity \dNdy and each of them fall on a single curve, implying that they depend only on the local density.

A nonzero cross-term \Rol coupling the ``out'' and ``long'' components is observed at low \kt in the forward direction of central events.
This demonstrates a breaking of the boost-invariance of the source function on the proton-going side.
In hydrodynamic models this indicates both longitudinal and transverse expansion, and indeed hydrodynamic predictions reproduce the rapidity dependence of \Rol in central (0--1\%) collisions.
This gives another indication of hydrodynamics that is only significant in central events.

The transverse HBT area $\detRt = \Rout \Rside - \Ros^2$ is proportional to the local multiplicity \dNdy at low \kt, suggesting a constant areal density.
At higher \kt ($\gtrsim 0.5 \GeV$) the volume element \detR is instead linear in the multiplicity.
The freeze-out volume increases steadily with \avgNpart.
With the standard Glauber geometry, the ratio of \detR to \Npart rises rapidly around $\avgNpart \gtrsim 12$, but in a \ac{GGCF} model the increase is much more modest.
A strict linear scaling of \detR with \avgNpart is not necessarily expected, but extreme deviations like that shown with the Glauber model are difficult to explain.
Previous results have supported the \ac{GGCF} models over standard Glauber, and these results provide some additional evidence for this view.
Even with allowances for fluctuations in the proton size, the expansion factor (i.e. ratio of final volume to initial size) still appears to be greater in central events, possibly indicating a turn-on of hydrodynamics.

The ratio of \Rout to \Rside is less than unity for all centrality and kinematic selections.
The value of $\Rout/\Rside$ is interpreted as growing with the lifetime of the source, since in hydrodynamic models the lifetime contributes to \Rout but not to \Rside.
It decreases with rising \kt, which is consistent with the description that higher-momentum particles freeze out from the source at earlier times.
The small value is consistent with an explosive expansion of the source.
There is little centrality dependence, although it is slightly higher in very central events, and there is no significant rapidity dependence of $\Rout/\Rside$.

The modulation of the HBT radii with respect to the second-order event plane in very central \pPb events with $\Nch \geq 150$ is also measured.
In events with a large flow the second-order Fourier components of the HBT radii can be extracted with good precision.
In events with low anisotropy the poor event plane resolution precludes distinguishing the modulation from zero.
A similar dependence of the radii on the azimuthal angle relative to the second-order event plane is observed as in \AA collisions.
The transverse radii are suppressed in-plane and enhanced out-of-plane, which is the same orientation required for the initial conditions in hydrodynamic models with a short lifetime of the medium.
The \kt-dependence of the Fourier components of the radii is also consistent with the predictions from hydrodynamics and shows that the ellipticity of the expanding source decreases throughout the duration of its evolution.
The longitudinal ratio \Rlong modulation indicates that the medium has a greater longitudinal expansion along the event plane.
These results support the interpretation that short-lived hydrodynamic evolution is a source of the flow-like azimuthal multiplicity distributions in central \pPb events and present a significant challenge to competing descriptions that do not directly link initial geometry to final-state momentum distributions.

In summary, the results presented in this dissertation provide detailed measurements of the \pPb source density as a function of all the most significant event-level and kinematic variables.
The highly-differential measurements include the first rapidity- and azimuthally-dependent femtoscopic results in \pA collisions.
Theoretical computations have already shown some success in describing central \pPb events and the opportunity to post-dict the Fourier components of the azimuthal HBT radii is available.
The evidence for hydrodynamic behavior is compelling in central \pPb; however, it remains incumbent upon the field to determine precisely how and where the onset of hydrodynamics occurs.

%% dijet asymmetry in ultra-central pPb??

Recent theoretical work has provided some explanation for the unreasonable success of hydrodynamics, even in systems that are not expected to last long enough to reach full thermal equilibrium.
Experiments should continue to work to establish a precise quantitative description of the domain of applicability of hydrodynamics.
This is challenging, since the turn-on is not likely to occur suddenly in any of the relevant physical variables like \Npart, multiplicity, and transverse momentum.
It now seems clear that central \pPb collisions evolve through hydrodynamics, but measurements of collective and hydrodynamic behavior need to be refined in peripheral \pPb and \pp collisions.
For the various correlation analyses used to probe the bulk behavior, the systematic effects in these smaller systems are much more significant due to lower \Nch, poorer event plane resolution, and the increased relative contribution of jets.

If the technique for constraining the jet contribution presented in this thesis is able to be improved in future iterations, femtoscopy would be able to better constrain the \kt-dependence of the source and determine where collective expansion is an appropriate description.
This may become increasingly feasible as fragmentation functions are better understood and the information is implemented into \mc event generators.

Azimuthal femtoscopy may be able to be extended down to mid-central \pPb, or even peripheral \pPb and \pp collisions, with a data sample a few orders of magnitude larger.
A larger dataset is necessary because the event plane resolution is poor outside of central \pPb collisions, and the correction procedure increases statistical uncertainties.
Computational demands for such a jump are likely prohibitive at the moment, but by the time Run 3 is underway at the \lhc, resources may be more readily available for such an approach.
